Finch: A Datastructure-Driven Array Programming Language

### 2024-03-19 main.tex
Data goes first
Update on Overleaf.

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics. Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more.  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}.  These frameworks have been the
subject of extensive industry investment to enable performant implementations,
and often operate at the peak capacity of the hardware they run on
\cite{lo_roofline_2015}.

The success and ubiquity of arrays is likely due to their simplicity. Since
their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. By \textbf{dense}, we mean that every entry the array
represents is stored faithfully at a location determined by a simple row or
column-major mapping from multidimensional space to linear memory. This
simplicity enables extensive interoperability, convenience layers, and
optimizations by breaking the abstraction barrier between array representation
and array storage.  Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectoriation,
loop-carry-dependencies, and more. Several optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.  However, while dense arrays are the easiest way to program, the
world is not all dense.

>>>>

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}.
%
These frameworks have been the
subject of extensive industry investment to enable performant implementations,
and often operate at the peak capacity of the hardware they run on
\cite{lo_roofline_2015}.

The success and ubiquity of arrays is likely due to their simplicity. 
%
Since
their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that every entry the array
represents is stored faithfully at a location determined by a simple row or
column-major mapping from multidimensional space to linear memory.
%
This
simplicity enables extensive interoperability, convenience layers, and
optimizations by breaking the abstraction barrier between array representation
and array storage.  
%
Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectoriation,
loop-carry-dependencies, and more. Several optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.  However, while dense arrays are the easiest way to program, the
world is not all dense.
